An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-07-07 cal_demo/ms.tex
First round addressing Judds comments

% Abstract of the paper
\begin{abstract}
We present the E-field Parallel Imaging Calibration (EPICal) algorithm, which addresses the 
need for a real-time calibration method for direct imaging radio astronomy correlators. Direct 
imaging involves a spatial fast Fourier transform of antenna \textbf{signals}, alleviating the harsh 
$\mathcal{O}(\Nant^2)$ computational scaling to a more gentle $\mathcal{O}(\Nant \log_2 
\Nant)$, which can save orders of magnitude in computation cost for next generation arrays 
consisting of hundreds to thousands of antennas. However, because signals are mixed in the 
correlator, gain correction must be applied on the front end. We develop the EPICal algorithm 
to form gain solutions in real time without ever forming visibilities. This method scales as the 
number of antennas, and produces results comparable to those from visibilities. Through 
simulations and application to Long Wavelength Array data we show this algorithm is a 
promising solution for next generation instruments.
\end{abstract}

@@

\textbf{The most common radio correlator designs of today, the FX and lag correlators \citep{rom99}},
cross-multiply the \textbf{signals} from all pairs of antennas, and 
the computation scales as the number of antennas squared, $\mathcal{O}(\Nant^2)$ 
\citep{bun04}. As the number of elements in future arrays grows, the computational cost will 
become prohibitively expensive, and exploring efficient correlator schemes is essential to 
enable next generation instruments \citep{lon00}. Meanwhile, radio transient monitoring 
requires access to high time and frequency resolution data to identify and characterize events 
such as fast radio bursts (FRBs, \citealt{lor07}), or to follow up gravitational wave candidates 
with radio observations \citep{abb16a,abb16b}. FRBs are highly unexplored at low frequencies 
(< 1 GHz), but are expected to occur on timescales $\Delta t \sim$ 1--10~ms \citep{tho13}. 
Recording the full visibility matrix for $\Nant \gtrsim 10^3$ arrays at this timescale leads to 
extremely high data write rates.

@@

A challenge common to all direct imaging algorithms is calibration of the antenna gains. With a 
traditional \textbf{cross correlator}, pair-wise visibilities are written to disk and used to calibrate 
\textbf{after observing before further processing} such as imaging. However, a direct imaging correlator mixes 
the signals from all antennas before averaging and writing to disk, making calibration a 
requirement at the front end, before imaging and averaging. Previous solutions have involved 
applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or 
integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve 
for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of 
FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

%%% New text
\textbf{We begin by considering the electric field vector incident on the ground as a function of time, 
$\mathbf{E}(\mathbf{r},t)$. The position vector, $\mathbf{r}$, is measured from some arbitrary 
origin. Following \citealt{cla99} we make a few assumptions in order to clarify notation and 
simplify the problem. First we note that the time ordered electric field can be expressed as a 
Fourier series with coefficients corresponding to frequencies $f$. This allows us to analyze the 
quasi-monochromatic components of the electric field, $\mathbf{E}(\mathbf{r},f,t)$, with the 
understanding that the full field is composed of all components and their corresponding Fourier 
functions. We retain the time dependence because in practice each coefficient will only be 
measured over some sampling period. }

\textbf{It is convenient to simplify notation by expressing the position vector in units of wavelengths, 
$\mathbf{r}\rightarrow \mathbf{r}/\lambda$. We will also assume all antennas lie in a plane, 
reducing the problem to two dimensions. While the MOFF algorithm in principle can account for 
non-coplanar arrays through the W-projection algorithm \citep{cor08}, the EPIC code does not 
currently support this so we leave it to future work.}

\textbf{Next we treat the electric field as a scalar quantity, ignoring polarization effects. This allows us 
to express the electric field as a scalar, $E(\mathbf{r},f,t)$, and treat subsequent field 
propagation and measurements as scalar rather than tensor multiplications. A full polarization 
treatment is required to construct polarized sky images, however the calibration will only 
require a sky model expressed in the instrumental polarization basis. We can therefore treat 
each polarization independently.}

@@

\textbf{Next we turn to the measurement performed by an antenna with far-field radiation pattern 
$\mathcal{W}_a(\hat{\mathbf{s}})$, where the subscript $a$ denotes a specific antenna. We 
adopt the common convention of normalizing the radiation pattern to have response of 1 at the 
center of the main lobe \citep[e.g.][]{nap99}. Consequently, the translation between the terminal 
voltages to electric field measurements has been absorbed in our definition of the radiation 
pattern. The ideal antenna electric field measurement can be expressed as 
\begin{equation}
E^I_a(f,t) = \int \mathcal{W}_a(\hat{\mathbf{s}})\mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega. 
\end{equation}
Equivalently we can express the effect of the antenna radiation pattern in the plane of the 
aperture.
\begin{equation}
E^I_a(f,t) = \int_{\mathrm{aperture}}W_a(\mathbf{r}-\mathbf{r}_a)E_a(\mathbf{r},f,t)\,d^2\mathbf{r}
\end{equation}
Here we used the Fourier relationship between the antenna's far-field radiation pattern and the 
aperture illumination pattern (sometimes referred to as the aperture distribution), $W_a(\mathbf{r}-\mathbf{r}_a)=\int  \mathcal{W}_a(\hat{\mathbf{s}}) \, d
\Omega$. 
While it is possible to define each antenna aperture illumination pattern in a way to absorb the 
relative position of the antenna, we keep the $\mathbf{r}_a$ coordinate to retain the separable
concepts of the primary beam (the product of two antenna radiation patterns) and the array
point spread function (square of the Fourier transform of the array layout).}

\textbf{The first step of the MOFF algorithm is to digitize the antenna signals and perform an FFT to 
channelize the time-ordered data. This is effectively estimating the quasi-monochromatic 
components of the electric field we already discussed. The sampling rate of the digitization 
determines the bandwidth of the observation, and the sampling period determines the minimum 
frequency channel resolution. The frequency Fourier transform operation is often referred to as 
the F-engine of a correlator. The MOFF shares the same F-engine design as an FX cross
correlator which performs the frequency transform (F-engine) before cross multiplying 
(X-engine). }

@@

The next step is to grid the \textbf{calibrated electric field measurements to a finite set of regularly spaced 
positions on the 2D plane of the ground, $\mathbf{r}_j$.
The extent of the grid is chosen to encompass the footprint of the antenna array.
MOFF uses the antenna aperture illumination patterns as the} gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. 
\textbf{Mathematically we can express the gridding operation as}
\begin{equation}\label{eq:gridding}
E'_j=\sum_a W_a(\mathbf{r}_j-\mathbf{r}_a) E'_a(t)
\end{equation}
\textbf{Formally each pixel is calculated by summing over all antennas and their aperture illumination patterns.
However the aperture illumination patterns are typically compact and non-overlapping, so most pixels
will only contain contribution from at most one antenna, allowing equation~\ref{eq:gridding}
to be calculated through a sparse matrix multiplication.}
This gridding 
achieves an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric field measurements directly. This 
operation must be performed for every time sample, before any averaging of the data.

@@

Finally, the images are squared and averaged in time to form \textbf{dirty images, 
$I'(\hat{\mathbf{s}}_k)$, then} written 
to disk. While operations have been reordered, the resulting image is equivalent to one 
produced by gridding and imaging visibilities. By shifting from cross-correlation of all pairs of 
antennas to a spatial FFT, the computational cost is significantly lowered for certain classes of 
compact arrays. The output bandwidth is also greatly reduced as visibilities are exchanged for 
gridded images. A full exploration of these scalings was presented in \citealt{thy15c}.

@@

As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it 
was suggested to form a correlation of the uncalibrated antenna measurements with the \textbf{complex conjugate of} an image 
pixel from the output of the correlator. This is a statistically stationary quantity, which can be 
related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum 
needed to calibrate a simple sky of a single point source. However, we aim for a more 
generalized solution for arbitrarily complex sky models. We therefore study the full expression 
for the antenna-pixel correlation,
\begin{equation}\label{eq:Kna_def}
\Kna \equiv \left<E_a^{(n)}(t) \mathcal{E}'^{*(n)}(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ 
calibration loop, 
\textbf{the superscript $^*$ indicates complex conjugation,}
and \spix\, is the pixel center nearest a bright calibrator of interest. The 
following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which 
contains a bright source to achieve a high signal to noise.

@@

\textbf{We will ultimately wish to model the right side of equation~\ref{eq:Kna} in order to solve for the
antenna gains. However, the self-correlated noise, $\left<|n_a(t)|^2\right>$, can often be 
significantly larger than the cross correlation power, and difficult to model precisely. It 
is therefore beneficial to subtract the self-correlation directly from $\Kna$. By properly
accounting for the various weighting and phasing terms, we can use self-correlations of the
uncalibrated antenna electric field measurements to remove the corresponding
terms from equation~\ref{eq:Kna}.}
\begin{align}\label{eq:Cna}
\Cna & \equiv \Kna - \frac{1}{\Nant g^{*(n)}_a}\mathcal{W}^*_a(\spix)e^{-2\pi i \spix \cdot \ra} \left<|E_a|^2\right>_t\nonumber \\
& = \frac{g_a}{\Nant} \sum_{b\ne a} \frac{g^*_b}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} V^I_{ab}
\end{align}

@@

We show schematically the process of calibrating a direct imaging correlator in 
Fig.~\ref{fig:schematic}. Computationally expensive steps that must be performed \textbf{in near real-time} 
are shown inside the gray box. The uncalibrated antenna signals are tapped out after the 
F-engine and correlated against the output image pixel of interest (i.e., equation~\ref{eq:Cna} is 
\textbf{computed in near real-time}). The correlated values are then \textbf{used to estimate the gains}
using equation~\ref{eq:cal_solution}, and additional fitting if desired. \textbf{This step does not have
the speed requirement of the steps in the gray box because the correlations have been 
averaged in time.}
The gains are then passed back 
to the correlator to update the calibration for subsequent integration intervals.

@@

We conclude this section by connecting our calibration expression to that found in a visibility 
framework. In the limit of a single bright calibrating source \textbf{at center,} we can greatly 
simplify equations~\ref{eq:Cna} and~\ref{eq:cal_solution}. We will assume the \textbf{radiation patterns} are 
normalized such that $\mathcal{W}(0)=1$. We can further drop the exponential phase terms 
because $\spix$ \textbf{is perpendicular to our planar antenna array.} We then absorb the true gains into the true visibilities in 
equation~\ref{eq:Cna} to express as a sum of measured, uncalibrated, visibilities.
\begin{equation}
\mathcal{C}^{(n)}_{a,0} \rightarrow \frac{1}{\Nant}\sum_{b\ne a} \frac{1}{g^{*(n)}_b} V_{ab}
\end{equation}

@@

For our unknown gains, we create a set of random complex numbers where the amplitude is 
drawn from an approximately gaussian distribution centered around 1 with width 0.25, and 
completely random phase. These are our ``true gains", and we apply them to the frequency
domain simulated antenna electric field \textbf{measurements} as in equation~\ref{eq:apply_gain}. Our analysis is 
blind to these values until the end of the process to check accuracy. The gain estimates are 
initialized with unity, $g^{(0)}_a=1$.

@@

After the gains have converged, we see both the phases and amplitudes to continue to 
fluctuate coherently. This is due to the stochastic fluctuations of the sources 
themselves, \textbf{which are simulated as zero mean random processes with variance proportional
to their flux density.}
For this simulation we restricted each calibration 
iteration to only 10~ms integration time. In principle a calibration implementation could use 
short integration times to allow the gains to converge, then increase the integration time to 
reduce this noise.

@@

We use a simulated sky consisting of a 5~Jy calibrator source, and 49 other random sources 
with apparent flux densities 0.2 to 0.5 Jy (total sky power $\approx$ 23 Jy). We generated 
these sources randomly, but kept them fixed for each run. 
\textbf{We simulate the same antenna layout as in Section~\ref{sec:sim}. 
However, we increase the number of frequency channels simulated to 64.}
Because our calibration loop treats 
frequency channels independently, each channel can be treated as a separate trial of the 
simulation, and is used to better estimate the statistics.

@@

For each simulation run we find gain solutions using four methods:
\begin{enumerate}[i.]
\item EPICal, using a full sky model
\item EPICal, using a single point source model
\item Visibility-based, using a full sky model
\item Visibility-based, using a single point source model.
\end{enumerate}
In the cases where we use the full sky model, the model visibilities, $V^M_{ab}$, used in the 
calibration loop are created using all point sources in the simulated sky. For the single point 
source model, we only allow the calibration loop to be privy to the bright calibrator source. 
While we vary the model used for calibration, the simulated ``true sky'' remains fixed with all 50 
point sources.

@@

When using the single point source model, all calibration methods in Fig.~\ref{fig:errors} trend 
downward until they reach an error floor due to the confusion sources that were not modeled. 
The EPICal and visibility-based solutions reach a similar floor, but EPICal achieves a 
marginally lower level (on average 5\% lower). This can be attributed to the same reason 
EPICal underperformed with the full sky model: EPICal only used a single pixel in the sky to 
form its solutions. When using the full sky model, it was down weighting the information in the 
rest of the sky. But when the sky model only contains the bright point source, the single pixel 
used in EPICal is the location where this model is most accurate, effectively down weighting 
the pixels with incomplete sky model. In the regime typical of HERA discussed above, EPICal's 
error floor was 10\% lower than that of visibility-based solutions.

@@

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, 
and 244 were omitted.} produced unstable gain solutions, and in fact corrupted the entire array. 
We therefore \textbf{omitted these antennas from our analysis,} resulting in a total of 248 antennas to 
calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and 
update our gain estimates over 10 \textbf{samples} (51.2 ms). We iterate the loop 30 times for a 
total of 1.536 seconds. The results of this calibration experiment are shown in Figs.~
\ref{fig:data_phase}~--~\ref{fig:data_images}.

Figure~\ref{fig:data_phase} shows the phase of our gain estimates over 30 calibration 
iterations, again with each colored line representing a different antenna. Given the quality of 
uncalibrated image demonstrated in \cite{thy15c}, it is a bit surprising to see the phase 
variation in our solutions. However, the phases are relatively flat after about 15 iterations 
(modulo noise), and exhibit a central ``trunk" where the majority of phases are congregated. 
This behavior is suggestive that while the uncalibrated data were able to produce a viable 
image, the minor changes from our solutions will focus the image and improve the quality. The 
actual location of the ``trunk" (slightly negative) is simply determined by the reference antenna 
chosen to have identically zero phase, but happens to be slightly more positive than the bulk of 
antennas. For plotting clarity, we unwrapped phases resulting in phases that appear to exceed 
$\pm \pi$.

@@

\textbf{To ensure the level of noise in our solutions is reasonable we compared to the fractional errors
from our simulations shown in Fig.~\ref{fig:errors}. We first use the Global Sky Model
\citep{deo08} to generate a mock observed sky at 74.03 MHz using the PyGSM python 
module\footnote{https://github.com/telegraphic/PyGSM} \citep{pri16}. We integrate this model
to estimate a total sky flux power of 171,000~Jy, meaning our calibration pixel containing
Cyg A accounts for  approximately $20,539/171,000 \approx 12\%$ of the total sky power. 
This places us in a regime very similar to the incomplete sky model simulations of the 
previous section, where the calibration pixel accounted for 21.7\% of the sky power. We 
should therefore expect our fractional gain noise levels to be at best around 0.1, the level 
to which the simulations asymptotically approached. }

@@

To compare the image qualities quantitatively we compute the dynamic range, defined as the 
peak of the image over the noise level. We estimate the noise level as the median of the 
absolute deviation of the image, a robust statistic. With this metric we find the \textbf{uncalibrated 
and} calibrated images to have \textbf{dynamic ranges of 88.5 and 137, respectively, a 55\% 
improvement.}

@@

\textbf{Improving sky model.} In this work we used a perfect sky model (for simulation), or a 
very simple sky model (for LWA data). In principle the images produced by the EPIC correlator 
can be used to improve the sky model used in calibration -- similar to a major loop in 
self-calibration. This can be especially useful for compact, widefield arrays which require a model of 
both compact and diffuse sources over a large patch of sky. In addition to requiring the sky 
model, the calibration also requires this model to be attenuated by the primary beam, which 
can be difficult to measure at the precision necessary \citep[e.g.][]{neb15,vir14,thy15b}. 
However, the direct imaging correlator provides exactly the model necessary in real time, and 
can be iterated over to improve the images and gain solutions.

@@

The work here will serve as a foundation for further development. We have shown that the 
EPICal algorithm produces reliable calibration solutions, and have identified several aspects to 
increase the scope. With next generation instruments in the planning and development stages, 
EPICal is poised to greatly contribute to their success. The software is integrated into the EPIC 
package and freely available to use in simulations or \textbf{post-processing} of data. Work is 
underway to port the code to real-time GPU systems for deployment.

>>>>

% Abstract of the paper
\begin{abstract}
We present the E-field Parallel Imaging Calibration (EPICal) algorithm, which addresses the 
need for a real-time calibration method for direct imaging radio astronomy correlators. Direct 
imaging involves a spatial fast Fourier transform of antenna \textbf{signals}, alleviating 
\textbf{an 
$\mathcal{O}(\Nant^2)$ computational bottleneck typical in radio correlators, and yielding a more gentle $\mathcal{O}(\Nant \log_2 
\Nant)$ scaling.} \textbf{This} can save orders of magnitude in computation cost for next generation arrays 
consisting of hundreds to thousands of antennas. However, because signals are mixed in the 
correlator, gain correction must be applied \textbf{prior to correlation, rather than on visibilities post-correlation.} We develop the EPICal algorithm 
to form gain solutions in real time without ever forming visibilities. This method scales as the 
number of antennas, and produces results comparable to those from visibilities. Through 
simulations and application to Long Wavelength Array data we show this algorithm is a 
promising solution for next generation instruments.
\end{abstract}

@@

\textbf{The most common radio correlator designs of today, the FX and lag correlators \citep{rom99}},
cross-multiply the \textbf{signals} from all pairs of \textbf{antennas.
This} computation scales as the number of antennas squared, $\mathcal{O}(\Nant^2)$ 
\citep{bun04}. As the number of elements in future arrays grows, the computational cost will 
become prohibitively expensive, and exploring efficient correlator schemes is essential to 
enable next generation instruments \citep{lon00}. Meanwhile, radio transient monitoring 
requires access to high time and frequency resolution data to identify and characterize events 
such as fast radio bursts (FRBs, \citealt{lor07}), or to follow up gravitational wave candidates 
with radio observations \citep{abb16a,abb16b}. FRBs are highly unexplored at low frequencies 
(< 1 GHz), but are expected to occur on timescales $\Delta t \sim$ 1--10~ms \citep{tho13}. 
Recording the full visibility matrix for $\Nant \gtrsim 10^3$ arrays at this timescale leads to 
extremely high data write rates.

@@

A challenge common to all direct imaging algorithms is calibration of the antenna gains. With a 
traditional \textbf{cross correlator}, pair-wise visibilities are written to disk and used to calibrate 
\textbf{after observing and before further processing} such as imaging. However, a direct imaging correlator mixes 
the signals from all antennas before averaging and writing to disk, making calibration a 
requirement at the front end, before imaging and averaging. Previous solutions have involved 
applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or 
integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve 
for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of 
FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

%%% New text
\textbf{We begin by considering the electric field vector incident on the ground as a function of time, 
$\mathbf{E}(\mathbf{r},t)$. The position vector, $\mathbf{r}$, is measured from some arbitrary 
origin. Following \citealt{cla99} we make a few assumptions in order to clarify notation and 
simplify the problem. First we note that the time ordered electric field can be expressed as a 
Fourier series with coefficients corresponding to frequencies $f$. This allows us to analyze the 
quasi-monochromatic components of the electric field, $\mathbf{E}(\mathbf{r},f,t)$, with the 
understanding that the full field is composed of all components and their corresponding Fourier 
functions. We retain the time dependence because in practice each Fourier component will only be 
measured over discrete sampling intervals. }

\textbf{It is convenient to simplify notation by expressing the position vector in units of wavelengths relative to the monochromatic component of interest, 
$\mathbf{r}\rightarrow \mathbf{r}/\lambda$. We will also assume all antennas lie in a plane, 
reducing the problem to two dimensions. While the MOFF algorithm in principle can account for 
non-coplanar arrays through the W-projection algorithm \citep{cor08}, the EPIC code does not 
currently support this so we leave it to future work.}

\textbf{Next we treat the electric field as a scalar quantity, ignoring polarization effects. This allows us 
to express the electric field as a scalar, $E(\mathbf{r},f,t)$, and treat subsequent field 
propagation and measurements as scalar rather than tensor multiplications. A full polarization 
treatment is required to construct polarized sky images, however the calibration will only 
require a sky model expressed in the instrumental polarization basis. We can therefore treat 
each instrument polarization independently.}

@@

\textbf{Next we turn to the measurement performed by an antenna with far-field radiation pattern 
$\mathcal{W}_a(\hat{\mathbf{s}})$, where the subscript $a$ denotes a specific antenna. We 
adopt the common convention of normalizing the radiation pattern to have response of unity at the 
center of the main lobe \citep[e.g.][]{nap99}. Consequently, the translation between the terminal 
voltages to electric field measurements has been absorbed in our definition of the radiation 
pattern. The ideal antenna electric field measurement can be expressed as 
\begin{equation}
E^I_a(f,t) = \int \mathcal{W}_a(\hat{\mathbf{s}})\mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega. 
\end{equation}
Equivalently we can express the effect of the antenna radiation pattern in the plane of the 
aperture.
\begin{equation}
E^I_a(f,t) = \int_{\mathrm{aperture}}W_a(\mathbf{r}-\mathbf{r}_a)E_a(\mathbf{r},f,t)\,d^2\mathbf{r}
\end{equation}
Here we used the Fourier relationship between the antenna's far-field radiation pattern and the 
aperture illumination pattern (sometimes referred to as the aperture distribution), 
\begin{equation}
W_a(\mathbf{r}-\mathbf{r}_a)=\int  \mathcal{W}_a(\hat{\mathbf{s}}) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d
\Omega.
\end{equation} 
While it is possible to define each antenna aperture illumination pattern in a way to absorb the 
relative position of the antenna, we keep the $\mathbf{r}_a$ coordinate to retain the separable
concepts of the primary beam (the product of two antenna radiation patterns) and the array
point spread function (square of the Fourier transform of the array layout).}

\textbf{The first step of the MOFF algorithm is to digitize the antenna signals and perform an FFT to 
channelize the time-ordered data. This is effectively estimating the quasi-monochromatic 
components of the electric field. The sampling rate of the digitization 
determines the bandwidth of the observation, and the sampling period determines the minimum 
frequency channel resolution. The frequency Fourier transform operation is often referred to as 
the F-engine of a correlator. The MOFF shares the same F-engine design as an FX cross
correlator which performs the frequency transform (F-engine) before cross multiplying 
(X-engine). }

@@

The next step is to grid the \textbf{calibrated electric field measurements to a finite set of regularly spaced 
positions on the plane of the ground, $\mathbf{r}_j$.
The extent of the grid is chosen to encompass the footprint of the antenna array.
MOFF uses the antenna aperture illumination patterns as the} gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. 
\textbf{Mathematically we can express the gridding operation as}
\begin{equation}\label{eq:gridding}
E'_j=\sum_a W_a(\mathbf{r}_j-\mathbf{r}_a) E'_a(t)
\end{equation}
\textbf{Formally each pixel is calculated by summing over all antennas and their aperture illumination patterns.
However the aperture illumination patterns are typically compact and non-overlapping, so most pixels
will only contain contribution from one antenna, allowing equation~\ref{eq:gridding}
to be calculated through a sparse matrix multiplication.}
This gridding 
achieves an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric field measurements directly. This 
operation must be performed for every time sample, before any averaging of the data.

@@

Finally, the images are squared and averaged in time to form \textbf{dirty images, 
$I'(\hat{\mathbf{s}}_k)$, then} written 
to disk. While operations have been reordered, the resulting image is equivalent to one 
produced by gridding and imaging visibilities. By shifting from cross-correlation of all pairs of 
antennas to a spatial FFT, the computational cost is significantly lowered for certain classes of 
compact arrays. The output bandwidth is also greatly reduced \textbf{in many scenarios} as visibilities are exchanged for 
gridded images. A full exploration of these scalings was presented in \citealt{thy15c}.

@@

As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it 
was suggested to form a correlation of the uncalibrated antenna measurements with the \textbf{complex conjugate of} an image 
pixel from the output of the correlator. This is a statistically stationary quantity, which can be 
related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum 
needed to calibrate \textbf{in the case of a single point source sky}. However, we aim for a more 
generalized solution for arbitrarily complex sky models. We therefore study the full expression 
for the antenna-pixel correlation,
\begin{equation}\label{eq:Kna_def}
\Kna \equiv \left<E_a^{(n)}(t) \mathcal{E}'^{*(n)}(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ 
calibration loop, 
\textbf{the superscript $^*$ indicates complex conjugation,}
and \spix\, is the pixel center nearest a bright calibrator of interest. The 
following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which 
contains a bright source to achieve a high signal to noise \textbf{ratio}.

@@

\textbf{We will ultimately wish to model the right side of equation~\ref{eq:Kna} in order to solve for the
antenna gains. However, the self-correlated noise, $\left<|n_a(t)|^2\right>$, can often be 
significantly larger than the cross correlation power, and difficult to model precisely. It 
is therefore beneficial to subtract the self-correlation directly from $\Kna$. By properly
accounting for the various weighting and phasing terms, we can use self-correlations of the
uncalibrated antenna electric field measurements to remove the corresponding
terms from equation~\ref{eq:Kna}:}
\begin{align}\label{eq:Cna}
\Cna & \equiv \Kna - \frac{1}{\Nant g^{*(n)}_a}\mathcal{W}^*_a(\spix)e^{-2\pi i \spix \cdot \ra} \left<|E_a|^2\right>_t\nonumber \\
& = \frac{g_a}{\Nant} \sum_{b\ne a} \frac{g^*_b}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} V^I_{ab}.
\end{align}

@@

We show schematically the process of calibrating a direct imaging correlator in 
Fig.~\ref{fig:schematic}. Computationally expensive steps that must be performed \textbf{in near real-time} 
are shown inside the gray box. The uncalibrated antenna signals are tapped out after the 
F-engine and correlated against the output image pixel of interest (i.e., equation~\ref{eq:Cna} is 
\textbf{computed in near real-time}). The correlated values are then \textbf{used to estimate the gains}
using equation~\ref{eq:cal_solution}, and additional fitting if desired. \textbf{This step does not have
the stringent cadence requirement of the steps in the gray box because the correlations have been 
averaged in time.}
The gains are then passed back 
to the correlator to update the calibration for subsequent integration intervals.

@@

We conclude this section by connecting our calibration expression to that found in a visibility 
framework. In the limit of a single bright calibrating source \textbf{at zenith,} we can greatly 
simplify equations~\ref{eq:Cna} and~\ref{eq:cal_solution}. We will assume the \textbf{radiation patterns} are 
normalized such that $\mathcal{W}(0)=1$. We can further drop the exponential phase terms 
because $\spix$ \textbf{is perpendicular to our planar antenna array.} We then absorb the true gains into the true visibilities in 
equation~\ref{eq:Cna} to express as a sum of measured, uncalibrated, visibilities.
\begin{equation}
\mathcal{C}^{(n)}_{a,0} \rightarrow \frac{1}{\Nant}\sum_{b\ne a} \frac{1}{g^{*(n)}_b} V_{ab}
\end{equation}

@@

For our unknown gains, we create a set of random complex numbers where the amplitude is 
drawn from an approximately gaussian distribution centered around \textbf{unity} with width 0.25, and 
completely random phase. These are our ``true gains", and we apply them to the frequency
domain simulated antenna electric field \textbf{measurements} as in equation~\ref{eq:apply_gain}. Our analysis is 
blind to these values until the end of the process to check accuracy. The gain estimates are 
initialized with unity, $g^{(0)}_a=1$.

@@

After the gains have converged, we see both the phases and amplitudes continue to 
fluctuate coherently. This is due to the stochastic fluctuations of the sources 
themselves, \textbf{which are simulated as zero mean random processes with variance proportional
to their flux density.}
For this simulation we restricted each calibration 
iteration to only 10~ms integration time. In principle a calibration implementation could use 
short integration times to allow the gains to converge, then increase the integration time to 
reduce this noise.

@@

We use a simulated sky consisting of a 5~Jy calibrator source, and 49 other random sources 
with apparent flux densities 0.2 to 0.5 Jy (total sky power $\approx$ 23 Jy). We generated 
these sources randomly, but kept them fixed for each run. 
\textbf{We simulate the same antenna layout as in Section~\ref{sec:sim}. 
However, we increase the number of frequency channels simulated to 64, while keeping the flux of all sources constant across frequency.}
Because our calibration loop treats 
frequency channels independently, each channel can be treated as a separate trial of the 
simulation, and is used to better estimate the statistics.

@@

For each simulation run we find gain solutions using four methods:
\begin{enumerate}[i.]
\item EPICal, using a full sky model
\item EPICal, using a single point source model
\item Visibility-based, using a full sky model
\item Visibility-based, using a single point source model.
\end{enumerate}
In the cases where we use the full sky model, the model visibilities, $V^M_{ab}$, used in the 
calibration loop are created using all point sources in the simulated sky. For the single point 
source model, we only \textbf{include the bright calibrator source in the model visibilities.}
While we vary the model used for calibration, the simulated ``true sky'' remains fixed with all 50 
point sources.

@@

When using the single point source model, all calibration methods in Fig.~\ref{fig:errors} trend 
downward until they reach an error floor due to the confusion sources that were not modeled. 
The EPICal and visibility-based solutions reach a similar floor, but EPICal achieves a 
marginally lower level (on average 5\% lower). This can be attributed to the same reason 
EPICal underperformed with the full sky model: EPICal only used a single pixel in the sky to 
form its solutions. When using the full sky model, it was \textbf{under utilizing} the information in the 
rest of the sky. But when the sky model only contains the bright point source, the single pixel 
used in EPICal is the location where this model is most accurate, effectively down weighting 
the pixels with incomplete sky model. In the regime typical of HERA discussed above, EPICal's 
error floor was 10\% lower than that of visibility-based solutions.

@@

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, 
and 244 were omitted.} produced unstable gain solutions, and in fact corrupted the entire array. 
We therefore \textbf{omitted these antennas from our analysis,} resulting in a total of 248 antennas to 
calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and 
update our gain estimates over 10 \textbf{samples} (51.2 ms). We iterate the loop 30 times for a 
total of 1.536 seconds of data processed. The results of this calibration experiment are shown in Figs.~
\ref{fig:data_phase}~--~\ref{fig:data_images}.

Figure~\ref{fig:data_phase} shows the phase of our gain estimates over 30 calibration 
iterations, again with each colored line representing a different antenna. Given the quality of 
uncalibrated image demonstrated in \cite{thy15c}, \textbf{we had anticipated that the phase
solutions would be small, hence the relatively large amplitudes of the phase solutions we
found was somewhat unexpected.}
However, the phases are relatively flat after about 15 iterations 
(modulo noise), and exhibit a central ``trunk" where the majority of phases are congregated. 
This behavior is suggestive that while the uncalibrated data were able to produce a viable 
image, the minor changes from our solutions will focus the image and improve the quality. The 
actual location of the ``trunk" (slightly negative) is simply determined by the reference antenna 
chosen to have identically zero phase, but happens to be slightly more positive than the bulk of 
antennas. For plotting clarity, we unwrapped phases resulting in phases that appear to exceed 
$\pm \pi$.

@@

\textbf{To ensure the level of noise in our solutions is reasonable we compared to the fractional errors
from our simulations shown in Fig.~\ref{fig:errors}. We first use the Global Sky Model
\citep{deo08} to generate a mock observed sky at 74.03 MHz using the PyGSM python 
module\footnote{https://github.com/telegraphic/PyGSM} \citep{pri16}. We integrate this model
to estimate a total sky flux power of 171,000~Jy, meaning our calibration pixel containing
Cyg A accounts for  approximately $20,539/171,000 \approx 12\%$ of the total sky power. 
This places us in a regime similar to the incomplete sky model simulations of the 
previous section, where the calibration pixel accounted for 21.7\% of the sky power. We 
should therefore expect our fractional gain noise levels to be at best around 0.1, the level 
to which the simulations asymptotically approached. }

@@

To compare the image qualities quantitatively we compute the dynamic range, defined as the 
peak of the image over the noise level. We estimate the noise level as the median of the 
absolute deviation of the image. With this metric we find the \textbf{uncalibrated 
and} calibrated images to have \textbf{dynamic ranges of 88.5 and 137, respectively, a 55\% 
improvement.}

@@

\textbf{Improving sky model.} In this work we used a perfect sky model (for simulation), or a 
very simple sky model (for LWA data). In principle the images produced by the EPIC correlator 
can be used to improve the sky model used in calibration -- similar to a major loop in 
self-calibration. This can be especially useful for compact, widefield arrays which require a model of 
both compact and diffuse sources over a large patch of sky. In addition to requiring the sky 
model, the calibration also requires this model to be attenuated by the primary beam, which 
can be difficult to measure at the precision necessary \citep[e.g.][]{neb15,vir14,thy15b}. 
However, the direct imaging correlator provides exactly the \textbf{product of the sky and beam} necessary in real time, and 
can be iterated over to improve the images and gain solutions, \textbf{leaving final beam
correction to post-processing as in traditional processing.}

@@

The work here will serve as a foundation for further development. We have shown that the 
EPICal algorithm produces reliable calibration solutions, and have identified several aspects to 
increase the scope. With next generation instruments in the planning and development stages, 
EPICal is poised to \textbf{enable new design spaces}. The software is integrated into the EPIC 
package and freely available to use in simulations or \textbf{post-processing} of data. Work is 
underway to port the code to real-time GPU systems for deployment.
